---
title: "Transcripomics"
output: 
  html_document: 
    fig_width: 10
    fig_height: 12
---

## Part 1 : Inference

```{r}
if (!require("BiocManager", quietly = TRUE)){
  install.packages("BiocManager")}

if (!require("factoextra", quietly = TRUE)){
  install.packages("factoextra")}

if (!require("DESeq2", quietly = TRUE)){
  BiocManager :: install("DESeq2")}

if (!require("apeglm", quietly = TRUE)){
  BiocManager :: install("apeglm")}

if (!require("EnhancedVolcano", quietly = TRUE)){
  BiocManager :: install("EnhancedVolcano")}

if (!require("vsn", quietly = TRUE)){
  BiocManager :: install("vsn")}

if (!require("hexbin", quietly = TRUE)){
  BiocManager :: install("hexbin")}

if (!require("caret", quietly = TRUE)){
  install.packages("caret")}

if (!require("glmnet", quietly = TRUE)){
  install.packages("glmnet")}

if (!require("mixOmics", quietly = TRUE)){
  BiocManager :: install("mixOmics")}


library(factoextra)
library(DESeq2)
library(MASS)
library(EnhancedVolcano)
library(vsn)
library(caret)
library(glmnet)
library(mixOmics)
```

```{r}
conditions = read.csv("Inf/E-ENAD-8-experiment-design.tsv",header = TRUE, sep = "\t")
expression = read.csv("Inf/E-ENAD-8-raw-counts.tsv",header = TRUE, sep = "\t")

# Dimension of the datasets
print("Dimensions of the conditions set")
dim(conditions)
print("Dimensions of the expression set")
dim(expression)

print("Sample count per experimental condition and cultivar")
table(conditions$Factor.Value.environmental.stress., conditions$Factor.Value.cultivar.)

# format Table
row.names(expression) = expression$Gene.ID
expression$Gene.ID = NULL
expression$Gene.Name = NULL
expression = round(expression)

# remove genes with 0 values
expression = expression[rowSums(expression != 0)> 0,]

# Log-transformed  matrix
log_expression = log(expression + 1)

# transpose expression matrix
expression_t = t(expression)

# Log-transformed transposed matrix
log_expression_t = t(log_expression)

# VST
model_vst = vsn2(as.matrix(expression))
vst_expression = predict(model_vst, as.matrix(expression))

```

```{r}
# plot the heteroskedasticity of the data
meanSdPlot(as.matrix(expression))
meanSdPlot(as.matrix(log_expression))
meanSdPlot(vst_expression)

```

```{r}
# Add condition label as factor
conditions$Factor.Value.environmental.exp_cond. = as.factor(paste(gsub(" ", "-",conditions$Factor.Value.cultivar), ifelse(conditions$Factor.Value.environmental.stress=="none","ctl","ost"), sep="_"))

conditions$Factor.Value.environmental.ost_cond. = ifelse(conditions$Factor.Value.environmental.stress=="none",0,1)

exp_condition_labels = conditions$Factor.Value.environmental.exp_cond.

ost_condition_labels = conditions$Factor.Value.environmental.ost_cond.


# PCA analysis for all conditions
pca_combined = prcomp(expression_t, scale. = TRUE, na.rm = TRUE)

 fviz_pca_ind(pca_combined,label= "none", habillage = exp_condition_labels, palette = c("blue","purple", "red", "orange","green", "yellow"), addEllipses = FALSE, title = "PCA of individuals per cultivar & controls vs osmotic stress - Non Normalized")
 
  fviz_pca_ind(pca_combined,label= "none", habillage = ost_condition_labels, palette = c("blue", "red"), addEllipses = TRUE, title = "PCA of individuals controls vs osmotic stress - Non Normalized")
  
 pca_log_expression = prcomp(log_expression_t, scale. = FALSE, na.rm =TRUE)

  fviz_pca_ind(pca_log_expression,label= "none", habillage = exp_condition_labels, palette = c("blue","purple", "red", "orange","green", "yellow"), addEllipses = FALSE, title = "PCA of individuals per cultivar & controls vs osmotic stress")
 
  fviz_pca_ind(pca_log_expression,label= "none", habillage = ost_condition_labels, palette = c("blue", "red"), addEllipses = TRUE, title = "PCA of individuals controls vs osmotic stress")
 
```

```{r}
# splitting data in subsets
cond_subset_Cachaco = conditions[conditions$Factor.Value.environmental.exp_cond.%in%c("Cachaco_ctl", "Cachaco_ost"),]
exp_sub_Cachaco = expression_t[cond_subset_Cachaco$Run,]

cond_subset_Grande_Naine = conditions[conditions$Factor.Value.environmental.exp_cond.%in%c("Grande-Naine_ctl", "Grande-Naine_ost"),]
exp_sub_Grande_Naine = expression_t[cond_subset_Grande_Naine$Run,]

cond_subset_Mbwazirume = conditions[conditions$Factor.Value.environmental.exp_cond.%in%c("Mbwazirume_ctl", "Mbwazirume_ost"),]
exp_sub_Mbwazirume = expression_t[cond_subset_Mbwazirume$Run,]


# PCA on banana subsets
pca_Cachaco = prcomp(exp_sub_Cachaco)
pca_Grande_Naine = prcomp(exp_sub_Grande_Naine)
pca_Mbwazirume = prcomp(exp_sub_Mbwazirume)

  fviz_pca_ind(pca_Cachaco,label= "none", habillage = cond_subset_Cachaco$Factor.Value.environmental.exp_cond., palette = c("blue", "red"), addEllipses = FALSE)

   fviz_pca_ind(pca_Grande_Naine,label= "none", habillage = cond_subset_Grande_Naine$Factor.Value.environmental.exp_cond., palette = c("blue", "red"), addEllipses = FALSE)

    fviz_pca_ind(pca_Mbwazirume,label= "none", habillage = cond_subset_Mbwazirume$Factor.Value.environmental.exp_cond., palette = c("blue", "red"), addEllipses = FALSE)

```

```{r}
gene_mean = rowMeans(expression+1)
gene_var = apply(expression+1, 1, var)

# log_gene_mean = log(gene_mean)
# log_gene_var = log(gene_var)
# 
# plot(log_gene_mean, log_gene_var, log="xy",xlab="Gene-Wise mean", ylab="Gene-Wise variance")
# abline(a=0,b=1, col="green")

df_meanvar <- data.frame(mean = gene_mean, var = gene_var)

ggplot(df_meanvar, aes(x = mean, y = var)) +
  geom_point(alpha = 0.4) +
  scale_x_log10() +
  scale_y_log10() +
  geom_abline(intercept = 0, slope = 1, color = "green") +  # Poisson
  geom_smooth(method = "loess", color = "red") +             # Trend
  labs(x = "Gene-Wise Mean", y = "Gene-Wise Variance", 
       title = "Mean-Variance Plot with Trend Line")

cor(log10(gene_mean), log10(gene_var), method = "pearson")
```

```{r}
dds = DESeqDataSetFromMatrix(countData = expression, colData = conditions, design = ~ Factor.Value.environmental.ost_cond. )

dds = DESeq(dds)

plotDispEsts(dds)

```

```{r}
#Default usage for generating results
results = results(dds)
results = results[order(results$log2FoldChange),]

#Volcano plot
EnhancedVolcano(results, x="log2FoldChange", y="padj",FCcutoff = 2, pCutoff = 0.05, lab=rownames(results), labSize = 2.0, ylim = c(0,30), xlim = c(-5,7), title = "Volcano Plot With default parameters")

#MA plot
plotMA(results, ylim=c(-5,5))

# results on p-value
results[which(results$pvalue < 0.05), ]
results[which(results$pvalue < 0.05 & results$log2FoldChange >0), ]
results[which(results$pvalue < 0.05 & results$log2FoldChange <0), ]

# results on adjusted p-value
results[which(results$padj < 0.05), ]
results[which(results$padj < 0.05 & results$log2FoldChange >0), ]
results[which(results$padj < 0.05 & results$log2FoldChange <0), ]


# Number of DEG with adj-p >0.05 & |LFC| > 2
nrow(results[which(results$padj < 0.05 & results$log2FoldChange > 2),])
nrow(results[which(results$padj < 0.05 & results$log2FoldChange < -2),])
```

```{r}
filtered_res = results(dds, lfcThreshold = 2)

filtered_res[which(filtered_res$pvalue < 0.05), ]
filtered_res[which(filtered_res$pvalue < 0.05 & filtered_res$log2FoldChange >0), ]
filtered_res[which(filtered_res$pvalue < 0.05 & filtered_res$log2FoldChange <0), ]

filtered_res[which(filtered_res$padj < 0.05), ]
filtered_res[which(filtered_res$padj < 0.05 & filtered_res$log2FoldChange >0), ]
filtered_res[which(filtered_res$padj < 0.05 & filtered_res$log2FoldChange <0), ]


#Volcano plot
EnhancedVolcano(filtered_res, x="log2FoldChange", y="padj", lab=rownames(filtered_res), pCutoff = 0.05, FCcutoff = 2, labSize = 2.0, xlim = c(-2,8), ylim = c(0,30), title = "Volcano Plot lfcThreshold")

#MA plot
plotMA(filtered_res)
```

```{r}
results_shrinkage = DESeq2::lfcShrink(dds,coef=2)
results_shrinkage[which(results_shrinkage$pvalue < 0.05),]
results_shrinkage[which(results_shrinkage$padj < 0.05), ]
results_shrinkage[which(results_shrinkage$padj < 0.05 & results_shrinkage$log2FoldChange > 0), ]


#MA plot
plotMA(results_shrinkage, ylim =c(-5,5))

#Volcano plot
EnhancedVolcano(results_shrinkage, x="log2FoldChange", y="padj", pCutoff = 0.05, lab=rownames(results_shrinkage), FCcutoff = 2, labSize = 2.0, xlim = c(-2,8), ylim = c(0,30), title = "Volcano Plot With penalization")

# Number of DEG with adj-p >0.05 & |LFC| > 2
nrow(results_shrinkage[which(results_shrinkage$padj < 0.05 & results_shrinkage$log2FoldChange > 2),])
nrow(results_shrinkage[which(results_shrinkage$padj < 0.05 & results_shrinkage$log2FoldChange < -2),])
```

```{r}
# Adaptation du modèle
dds_adj = DESeqDataSetFromMatrix(countData = expression, colData = conditions, design = ~ Factor.Value.environmental.ost_cond. + Factor.Value.cultivar.)

dds_adj = DESeq(dds_adj)

plotDispEsts(dds_adj)

# results_adj = results(dds_adj)
results_adj = DESeq2::lfcShrink(dds,coef=2)

#Volcano plot
EnhancedVolcano(results_adj, x="log2FoldChange", y="padj", pCutoff = 0.05, lab=rownames(results_adj), FCcutoff = 2, labSize = 2.0, xlim = c(-2,8), ylim = c(0,30), title = "Volcano Plot for adjusted model")
# plot(results$log2FoldChange, -log10(results$pvalue))

#MA plot
plotMA(results_adj, ylim=c(-5,5))


results_adj[which(results_adj$pvalue < 0.05), ]
results_adj[which(results_adj$padj < 0.05), ]
results_adj[which(results_adj$padj < 0.05 & results_adj$log2FoldChange >0), ]
results_adj[which(results_adj$padj < 0.05 & results_adj$log2FoldChange <0), ]

# Number of DEG with adj-p >0.05 & |LFC| > 2
nrow(results_adj[which(results_adj$padj < 0.05 & results_adj$log2FoldChange > 2),])
nrow(results_adj[which(results_adj$padj < 0.05 & results_adj$log2FoldChange < -2),])

```

```{r}

```

## Part 2: Prediction

```{r}
# Functions for Lasso Stability Selection

#' A function for performing LASSO with stability selection
#'@param  X a matrix of predictors with a column corresponding to the foldid
#'@param y the vector of responses 
#'@param lambda a vector, a sequence for evaluating lambda
#'@param nboot an integer corresponding to the number of bootstraps
#'@param random.seed an integer, for replication
#'@param family corresponding to the type of model performed
#'@return a LASSO object
#'
LASSO.SS = function(X, y, lambda=NULL, nboot=100, random.seed=1234, family="gaussian"){
  
  coefs.boots = list()
  
  X = as.data.frame(X)
  
  coltype=apply(X,2, class)

  character.col = which(is.character(coltype))

  X = as.matrix(X[, -character.col])
  
  n=nrow(X); p=ncol(X)
  print(n);print(p)
  for(i in 1:nboot){
    # random.index = sample(1:n, 0.5*n, replace = FALSE)
    random.index = createDataPartition(y, p = 0.5, list = FALSE)
    
    X.sub = X[random.index,]
    y.sub = y[random.index]
    
    #Randomized LASSO
    scaling_factors = runif(p, 0.2,0.8)
    X.random = X.sub*scaling_factors
    
    if(is.null(lambda)) fit = glmnet(X.random, y.sub, alpha = 1, lambda=seq(0.1,0.001, by=-0.0005), family = family,standardize = FALSE) 
    else fit = glmnet(X.random, y.sub, alpha = 1, lambda=lambda, family = family,standardize = FALSE) 
    #lambda sequence is specified by hand 
    
    if(family == "multinomial"){
      coef_list = coef(fit)
      coefs = as.matrix(Reduce("+", coef_list))
      coefs = coefs[-1,]
    }
    else{
    coefs=as.matrix(coef(fit))[-1,]
    }
    
    coefs[coefs!=0] = 1
    
    coefs.boots[[i]]=coefs
  }
  
  
  stability.scores = Reduce("+", coefs.boots)/nboot
  #max.by.features = rowMax(stability.scores)
  
  
  return(stability.scores)
  
  
}


#'  A function for plotting stability path from the LASSO.SS function
#'  @param stability.scores a matrix as provided by LASSO.SS
#'  @param threshold a value between 0 and 1 corresponding to the threshold of the probability that the feature is kept across bootstraps
#'  @return a stability path plot
plot.LASSO.SS = function(stability.scores ,threshold=0.8){
  
  
  plot(1, type="n",ylim=c(0,1), xlim=c(1,ncol(stability.scores)) ,xlab="lambda index", ylab="Probability Selection")
  
  
  max.by.features = apply(stability.scores,1,max)
  
  for(i in 1:nrow(stability.scores)) {
    
    if(max.by.features[i] > threshold) lines(1:ncol(stability.scores),stability.scores[i,], col="red", lty=2) 
    else lines(1:ncol(stability.scores),stability.scores[i,])
    
  }
  abline(h=threshold, lwd=2, col="red")
}
```

```{r}
# Importing the data
pred_conditions = read.csv("Pred/E-MTAB-4045-experiment-design.tsv",header = TRUE, sep = "\t")
  
  
pred_expression = read.csv("Pred/E-MTAB-4045-query-results.tpms.tsv",header = FALSE, sep = "\t")


# Formating tables  
colnames(pred_expression) = pred_expression[5,]
rownames_pred_expression = pred_expression[-1:-5,1]
pred_expression =pred_expression[-1:-5,c(-1,-2)]
pred_expression = data.frame(lapply(pred_expression[,-1], as.numeric))
rownames(pred_expression) = rownames_pred_expression



dim(pred_expression)
dim(pred_conditions)

# remove genes with missing values
pred_expression = pred_expression[complete.cases(pred_expression),]

# Deprecated Used for testing purpose
# # Remove unique condition accross development stages
# pred_expression = pred_expression[,!colnames(pred_expression) %in% c("early.maturation.stage..axis.epidermis","early.maturation.stage..axis.parenchyma","early.maturation.stage..axis.plumule","early.maturation.stage..axis.shoot.apical.meristem","early.maturation.stage..axis.stele","early.maturation.stage..axis.vasculature","early.maturation.stage..cotyledon.abaxial.epidermis","early.maturation.stage..cotyledon.abaxial.parenchyma","early.maturation.stage..cotyledon.adaxial.epidermis","early.maturation.stage..cotyledon.adaxial.parenchyma","early.maturation.stage..cotyledon.vasculature","cotyledon.stage..embryo.proper.axis","cotyledon.stage..embryo.proper.cotyledon","early.maturation.stage..root.tip","early.maturation.stage..seed.coat.hourglass","early.maturation.stage..seed.coat.palisade","early.maturation.stage..seed.coat.parenchyma")]

# View the remaining genes in the dataset
dim(pred_expression)

# Log-transformed  matrix
log_pred_expression = log(pred_expression + 1)

# transpose expression matrix
pred_expression_t = t(pred_expression)

# Log-transformed transposed matrix
log_pred_expression_t = t(log_pred_expression)

# VST
model_vst_pred = vsn2(as.matrix(pred_expression))
vst_pred_expression = predict(model_vst_pred, as.matrix(pred_expression))

```

```{r}
# plot the heteroskedasticity of the data
meanSdPlot(as.matrix(pred_expression))
meanSdPlot(as.matrix(log_pred_expression))
meanSdPlot(vst_pred_expression)

```

Normalization tested showed improvement over non-normalized data, both VST and Log-transformation gave good results, we will opt for log-transformation as it's more computer efficient and allow better interpretation.

```{r}
pca_factomineR_Pred_expression = FactoMineR::PCA(log_pred_expression_t, scale.unit = FALSE)
```

```{r}
# Add developemental stage label as factor

developmental_stages <- sub("\\.{2}.*", "", colnames(pred_expression))
pred_condition_labels <- factor(developmental_stages)


# PCA analysis for genes
pca_pred_genes = prcomp(log_pred_expression, scale. = FALSE)

 fviz_pca_ind(pca_pred_genes,label= "none", addEllipses = TRUE)
 fviz_pca_var(pca_pred_genes,label= "none", addEllipses = TRUE)
 
# PCA analysis for individuals
pca_pred_indiv = prcomp(log_pred_expression_t, scale. = FALSE)

 fviz_pca_ind(pca_pred_indiv,label= "none", habillage = pred_condition_labels, addEllipses = TRUE)
 fviz_pca_var(pca_pred_indiv,label= "none", addEllipses = TRUE)

```

```{r}
k_mean_cluster_gene = kmeans(log_pred_expression,centers = 2, algorithm = "Hartigan-Wong")

fviz_cluster(k_mean_cluster_gene,log_pred_expression)



k_mean_cluster_indiv = kmeans(log_pred_expression_t,centers = 20, algorithm = "Hartigan-Wong")

fviz_cluster(k_mean_cluster_indiv,log_pred_expression_t)
```

Clustering with both PCA and K-means methods show that there is no clear differentiation between developmental stages but instead k-mean clustering showed a tendency to cluster per organism part.

```{r}
# defining training indexes
train_index <- createDataPartition(pred_condition_labels, p = 0.8, list = FALSE)
pred_expression_train = t(pred_expression[,train_index])
pred_expression_test = t(pred_expression[,-train_index])
pred_labels_train = pred_condition_labels[train_index]
pred_labels_test = pred_condition_labels[-train_index]

binomial_outcome = ifelse(pred_condition_labels %in% c("globular.stage","heart.stage"),1,0)
binomial_outcome_train = binomial_outcome[train_index]
binomial_outcome_test = binomial_outcome[-train_index]
```

```{r}
# Plot the regularization plot for the multinomial model
glmnet_pred_multinomial = glmnet(pred_expression_train, pred_labels_train,family="multinomial") #for binary family="binomial"
plot(glmnet_pred_multinomial, xvar = "lambda") #regularization plot

```

```{r}
# Multinomial LASSO
cv_glmnet_pred_multinomial = cv.glmnet(pred_expression_train, pred_labels_train,family="multinomial")
plot(cv_glmnet_pred_multinomial, xvar = "lambda") #regularization plot


# Predictions Mutinomial 
reponse_multi = stats::predict(cv_glmnet_pred_multinomial, pred_expression_test,type="class")
table(pred_labels_test,reponse_multi)

# Binomial LASSO
cv_glmnet_pred_Binomial = cv.glmnet(pred_expression_train, binomial_outcome_train,family="multinomial")
plot(cv_glmnet_pred_Binomial, xvar = "lambda") #regularization plot


# Predictions Binomial 
reponse_bino = stats::predict(cv_glmnet_pred_Binomial, pred_expression_test,type="class")
table(binomial_outcome_test,reponse_bino)

```

```{r}
#Stability selection Binomial
LASSO_SS_pred_binom = LASSO.SS(pred_expression_train, binomial_outcome_train, family = "binomial")
plot.LASSO.SS(LASSO_SS_pred_binom, threshold = 0.5) #stability path
```

```{r}
max_LASSO_SS_binom = apply(LASSO_SS_pred_binom, 1, max)
gene_selection = pred_expression[max_LASSO_SS_binom >= 0.5,]
gene_selection_binom_train = pred_expression_train[,max_LASSO_SS_binom >= 0.5]
gene_selection_binom_test = pred_expression_test[, max_LASSO_SS_binom >= 0.5]
```

```{r}

#Stability selection multinomial
LASSO_SS_pred_multi = LASSO.SS(pred_expression_train, pred_labels_train, family = "multinomial")
plot.LASSO.SS(LASSO_SS_pred_multi, threshold = 0.5) #stability path

```

```{r}
max_LASSO_SS_multi = apply(LASSO_SS_pred_multi, 1, max)
gene_selection_multi_train = pred_expression_train[,max_LASSO_SS_multi >= 0.5]
gene_selection_multi_test = pred_expression_test[,max_LASSO_SS_multi >= 0.5]
```

```{r}
# determining genes selected in both models
setdiff(colnames(gene_selection_multi_train),colnames(gene_selection_binom_train))
```

```{r}

# Partial Least Square Multinomial
PLS_DA_multi_pred = plsda(pred_expression_train,pred_labels_train, ncomp=4)
plotIndiv(PLS_DA_multi_pred)
plotVar(PLS_DA_multi_pred)

predict_multi_plsda = stats::predict(PLS_DA_multi_pred, pred_expression_test)
res_predict_multi_plsda <- predict_multi_plsda$class$max.dist[,4]
table(factor(res_predict_multi_plsda, levels = levels(pred_labels_train)), pred_labels_test)


```

```{r}
# Sparse model (with less features)

keepX_multi  = rep(10,4)

# Sparse Partial Least Square Multinomial
PLS_DA_multi_pred = splsda(pred_expression_train,pred_labels_train, ncomp=4, keepX = keepX_multi)
plotVar(PLS_DA_multi_pred)

predict_multi_splsda = stats::predict(PLS_DA_multi_pred, pred_expression_test)
res_predict_multi_splsda <- predict_multi_splsda$class$max.dist[,4]
table(factor(res_predict_multi_splsda, levels = levels(pred_labels_train)), pred_labels_test)


# Tuning Sparse Partial Least Square Multinomial
tune_PLS_DA_multi_pred = tune.splsda(pred_expression_train,pred_labels_train, ncomp=10, folds = 5, nrepeat = 5, measure = "BER", test.keepX = seq(10,100,10), progressBar = TRUE)

optimalNComp = tune_PLS_DA_multi_pred$choice.ncomp$ncomp
optimalKeepX =  tune_PLS_DA_multi_pred$choice.keepX

tune_PLS_DA_multi_pred = splsda(pred_expression_train,pred_labels_train, ncomp=optimalNComp, keepX =optimalKeepX)

plotVar(tune_PLS_DA_multi_pred)

predict_multi_splsda_tuned = stats::predict(tune_PLS_DA_multi_pred, pred_expression_test)
res_tuned_multi_splsda <- predict_multi_splsda_tuned$class$max.dist[,optimalNComp]
table(factor(res_tuned_multi_splsda, levels = levels(pred_labels_train)), pred_labels_test)

```
